# Regression and model validation
Week 2

The data used is from a study which investigates the relationship between teaching, studying habits and results of the students.

First we import the data which we wrangled and saved locally in the first part of the exercise. The data is set to a variable ```lrn14``` using read.table. 

```{r}
lrn14 <- read.table("C:/Users/anton.jyrkiainen/Documents/IODS-project/learning2014.txt")
```

We also need to import few dependancies for later use:
```{r}
library(GGally)
library(ggplot2)
library(dplyr)
```

By using str() we can have a look at the structure of the data. The output gives us meta information about the data in general and about each column itself: amount of rows and colums, name of the variables, data type and the first values.

```{r}
str(lrn14)
```

Using ```dim()``` we get an output which gives as the dimensions of the dataset.
```{r}
dim(lrn14)
```

Using this dataset we generate a graphical overview:
```{r}
p <- ggpairs(lrn14, mapping = aes(), lower = list(combo = wrap("facethist", bins = 20)))
p
```

And provide a summary:

```{r}
summary(lrn14)
```